---
title: "Excess deaths in Europe 2020--2021"
author: 
  - name: "Tomasz Przechlewski"
    email:  t.plata-przechlewski@psw.kwidzyn.edu.pl
    affiliation: "Powiślańska Szkoła Wyższa (Kwidzyn/Poland)"
description: (c) Tomasz Przechlewski / CC-BY license 
date: "2021"
output:
  html_document:
    includes:
      before_body: misc_hdr.html
---


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = F, message = F)
library("ggplot2")
library("dplyr")
library("scales")
library("ggthemes")
library("ggpubr")
library("tidyr")
library("readr")
library("stringr")
library("ISOweek")
library("knitr")

source <- "© NI-KW @ github.com/knsm-psw/ | source: Eurostat demo_r_mwk_ts (Deaths by week and sex)"
note <- source
mainColor <- "deeppink"
loessColor <- "steelblue"
### picture dimensions
pWd <- 10
pHt <- 8.5
pHtx <- 12.5
pWdx <- 15.0
```

## Total excess mortality

### The Data 

Deaths data from Eurostat (`demo_r_mwk_ts`):

```{r}
library("eurostat")
z <- get_eurostat("demo_r_mwk_ts",  stringsAsFactors = FALSE) %>%
  mutate (time = as.character(time)) %>%
  mutate (year = as.numeric(substr(time, 1, 4)), 
          week = as.numeric(substr(time, 6,7)),
          sex = as.factor(sex),
          geo = as.factor(geo)) %>%
  select (sex, geo, year, week, value=values) %>%
  filter (sex == 'T')

## country names
nn <- read.csv("ecdc_countries_names_codes3166.csv", sep = ';',  header=T, na.string="NA" )
```

We define (weekly) excess mortality as a difference between number of deaths in 2020--2021 (`z1`) and 
5 year average (2015--2019) for corresponding weeks (`z0`). The **absolute excess mortality**
(`exm`) is defined as `z1 - z0` and the **relative excess mortality** (`exp`) is
`(z1 - z0)/z0 · 100%`:

```{r, echo=T}
## mean weekly deaths 2015--2019
z0 <- z %>% filter ( year >= 2015 & year < 2020) %>% 
  group_by(geo, week) %>% 
  summarise (d0 = mean(value, na.rm=T))
## weekly deaths 2020--2021
z1 <- z %>% filter ( year > 2019 ) %>% 
  group_by(geo, year, week) %>% 
  summarise ( d1 = sum(value))

## join z0 z1 and compute differences
zz <- left_join(z0, z1, by=c("week", "geo")) %>% 
  drop_na(d0,d1) %>%
  mutate (exp = (d1 - d0)/d0 * 100, exm = d1 - d0 )
zz <- left_join(zz, nn, by=c('geo'='id'))
```

Compute last week each country reported:

```{r, echo=T}
## if NA then the country stop reporting in 2020
zz.last.week <- zz %>% filter (year == 2021) %>% 
  group_by (geo) %>% 
  summarise (lw = last(week)) %>%
    drop_na(lw)
latestweek <- max (zz.last.week$lw)
```

The latest data is from `r latestweek` week, the oldest data
is from `r min (zz.last.week$lw)` week. Countries which did not report 
in 2021 (UK for example, were removed)

We summarise all weeks (`exm`) for each country. Countries which 
reported last earlier than for `r latestweek -6` week are excluded.

```{r}
### total exmort 
zzt <- zz %>% group_by(geo) %>% summarise(country=last(country), 
                                          exm = sum(exm), nm = sum(d0)) %>%
  mutate (exmp = exm/nm * 100)

zztt <- left_join(zzt, zz.last.week, by='geo') %>%  drop_na(lw) %>%
  filter (lw >= latestweek - 6)
countries.left <- nrow(zztt)
```

There are `r countries.left` countries which reported last for week
`r latestweek -6` or later.


### Total excess mortality by country


```{r, warning=F, message=F, fig.width=pWd}
x.note <- sprintf("%s\nFirst column in red = last week reported (excluded countries that reported in %i week or earlier)", source, latestweek - 6 - 1 )

p5 <- zztt %>% ggplot(aes(x = reorder(country, exmp ))) +
  geom_bar(aes(y = exmp), stat="identity", alpha=.4, fill=loessColor ) +
  geom_text(aes(label=sprintf("%.1f", exmp), y= exmp ), 
            vjust=0.25, hjust=1.25, size=2, color='black' ) +
  geom_text(aes(label=sprintf("%.0f", lw), y= 0 ), 
            vjust=0.25, hjust=-1.25, size=2, color='brown4' ) +
  xlab(label="") +
  ylab(label="") +
  ggtitle("Total Excessive deaths in Europe 2020--2021",
          subtitle ="Sum of (deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  theme_nikw() +
  coord_flip() +
  labs(caption=x.note)
p5
ggsave(plot=p5, "EU_exdeaths_total.png", width=pWd)
```

The same as above with Poland highlighted

```{r, warning=F, message=F, fig.width=pWd}
p6 <- zztt %>%  mutate( base=ifelse(geo=='PL', "1", "0")) %>% 
  ggplot(aes(x = reorder(country, exmp ), fill=as.factor(base))) +
  geom_bar(aes(y = exmp), stat="identity", alpha=.4 ) +
  geom_text(aes(label=sprintf("%.1f", exmp), y= exmp ), 
            vjust=0.25, hjust=1.25, size=2, color='black' ) +
  geom_text(aes(label=sprintf("%.0f", lw), y= 0 ), 
            vjust=0.25, hjust=-1.25, size=2, color='brown4' ) +
  xlab(label="") +
  ylab(label="") +
  ggtitle("Total Excessive deaths in Europe 2020--2021",
          subtitle ="Sum of (deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  theme_nikw() +
  coord_flip() +
  scale_fill_manual( values = c( "1"="red", "0"="steelblue" ), guide = FALSE ) +
  labs(caption=x.note)
p6
ggsave(plot=p6, "EU_exdeaths_total_p.png", width=pWd)
```

### Dynamics of excess mortality 2020--2021

Faceted line plots

```{r, warning=F, fig.height=pHtx, fig.width=pWdx }
#zznn <- left_join(zz, nn, by=c('geo'='id'))
zznn <- zz

p7 <- zznn %>% filter (! geo %in% c('AL', 'AM', 'IE', 'IS', 'UK', 'CY', 'GE', 'LI', 'ME')) %>%
  mutate(date= as.Date(sprintf("%i-%i-1", year, week), "%Y-%U-%u") ) %>%
  ##ggplot(aes(x = date, y =exp, group=geo, color=geo )) +
  ggplot(aes(x = date, y =exp)) +
  facet_wrap( ~country, scales = "fixed", ncol = 4, shrink = F) +
  geom_point(size=.4) +
  geom_smooth(method="loess", size=1, se = F, color="red",  span =.25) +
  geom_hline(yintercept = 50, color="firebrick", alpha=.2, size=1) +
  scale_y_continuous(breaks=seq(-100, 200, by=50)) +
  coord_cartesian(ylim = c(-100, 200)) +
  ggtitle("Excessive deaths in Europe 2020--2021",
          subtitle ="(deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  theme_nikw() +
  xlab(label="%") +
  labs(caption=note)
p7
ggsave(plot=p7, "EU_exdeaths_total_x.png", width=pWd, height=pHtx)
```

Poland and the rest of Europe:

```{r, warning=F, fig.width=pWd}
p8 <- zznn %>% filter (! geo %in% c('AL', 'AM', 'IE', 'IS', 'UK', 'CY', 'GE', 'LI', 'ME')) %>%
  mutate( base=ifelse(geo=='PL', "1", "0")) %>% 
  mutate(date= as.Date(sprintf("%i-%i-1", year, week), "%Y-%U-%u") ) %>%
  ggplot(aes(x = date, y =exp,  color=as.factor(base ))) +
  ##ggplot(aes(x = date, y =exp)) +
  geom_point(size=.4) +
  geom_smooth(aes(x = date, y =exp, group=geo), method="loess", size=.3, se = F, span =.25) +
  geom_hline(yintercept = 50, color="firebrick", alpha=.2, size=1) +
  coord_cartesian(ylim = c(-100, 200)) +
  scale_color_manual( values = c( "1"="red", "0"="steelblue" ), guide = FALSE ) +
  theme_nikw() +
  ylab(label="%") +
  ggtitle("Excessive deaths in Europe 2020--2021 and Poland (red)",
          subtitle ="(deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  scale_y_continuous(breaks=seq(-100, 200, by=50)) +
  labs(caption=note)
p8
ggsave(plot=p8, "EU_exdeaths_total_y.png", width=pWd, height=pHt)
```

Polar coordinates (I like Nightingale rose chart)

```{r, warning=F,  fig.height=18}
p9 <- zznn %>% filter (! geo %in% c('AL', 'AM', 'IE', 'IS', 'UK', 'CY', 'GE', 'LI', 'ME')) %>%
  #mutate(date= as.Date(sprintf("%i-%i-1", year, week), "%Y-%U-%u") ) %>%
  #ggplot(aes(x = date, y =exp )) +
  ggplot(aes(x = week, y = exp, group=as.factor(year), color=as.factor(year) )) +
  geom_point(size=.25 ) +
  geom_line() +
  facet_wrap( ~country, scales = "fixed", ncol = 4, shrink = F) +
  geom_hline(yintercept = 50, color="orange", alpha=.25, size=1.5) +
  geom_hline(yintercept = 0, color='seagreen1', alpha=.25, size=1.5) +
  ##coord_cartesian(ylim = c(-100, 200)) +
  #scale_color_manual( values = c( "1"="red", "0"="steelblue" ), guide = FALSE ) +
  theme_nikw() +
  coord_polar(start=3*pi/2) +
  labs(caption="", color = "Year") +
  ggtitle("Excessive deaths in Europe 2020--2021",
          subtitle ="(deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  scale_color_manual(values=c("red3", "purple3", "darkgoldenrod3")) +
  scale_y_continuous(breaks=seq(-100, 200, by=50))

p9
ggsave(plot=p9, "EU_exdeaths_total_fn.png", width=pWd, height=pHtx)
```

## Excess mortality by age groups

### The Data

Deaths data from Eurostat (`demo_r_mwk_10_ts`; Deaths 
by week, sex and 10-year age group; 
https://ec.europa.eu/eurostat/databrowser/view/demo_r_mwk_10/default/table?lang=en):

```{r}
z <- get_eurostat("demo_r_mwk_10",  stringsAsFactors = FALSE) %>%
  mutate (time = as.character(time)) %>%
  mutate (year = as.numeric(substr(time, 1, 4)), 
          week = as.numeric(substr(time, 6, 7)),
          age = as.factor(age),
          sex = as.factor(sex),
          geo = as.factor(geo)) %>%
  select (age, sex, geo, year, week, value=values) %>%
  filter (sex == 'T')

levels(z$age)
```

The `age` groups: `r levels(z$age)`.
We skip all `sex` levels (`r levels(z$sex)`)  except `T` (ie total).
We combine `Y_LT10`--`Y50-59`
into one group (`00-59`) and recode the names of the remaining groups.
Group `Y_GE80` is redundant so we skip it:


```{r}
agegrps <- c('00-59', '60-69', '70-79' , '80-89', '90-and-more', 'total')

z <- z %>% select (age, sex, geo, year, week, value) %>%  
  filter (sex == 'T') %>% 
  filter ( age != 'Y_GE80' ) %>%
  mutate(age=recode(age, 
                    'Y_LT10' = '00-59',
                    'Y10-19' = '00-59',
                    'Y20-29' = '00-59',
                    'Y30-39' = '00-59',
                    'Y40-49' = '00-59',
                    'Y50-59' = '00-59',
                    'Y60-69' = "60-69",
                    'Y70-79' = "70-79",
                    'Y80-89' = "80-89",
                    'Y_GE90' = "90-and-more",
                    'TOTAL'  = "total")
  ) %>% 
  ## recalculate values
  group_by(age, sex, geo, year, week) %>%
  summarise(value = sum(value)) %>%
  as.data.frame()
```

There are `r `levels(z$age)` age groups now. 

### Veryfing data consistency

Check how big is `UNK` (unknown) group:

```{r}
## Sprawdzenie age== UNK
z.unk <- z %>% filter(age == 'UNK') %>%
  group_by(geo) %>%
  summarise(d = sum(value, na.rm = T))
total.unknown <- sum(z.unk$d)
```

There are `r `total.unknown` unclassified deaths. We can safely skip
this age group.

We compare total deaths and sum of the deaths from other age categories. There
should be no difference if the data is consistent:

```{r}
## wszystkie wg wieku (za wyj UNK)
z.kn <- z %>% filter (age != 'UNK') %>%
  group_by(geo, age) %>% summarise(d = sum(value, na.rm = T))

## suma wg 'total' i po kategoriach
z.kn.t <- z.kn %>% filter (age == 'total') %>% 
  group_by(geo) %>% summarise(td = sum(d))
z.kn.mf <- z.kn %>% filter (age != 'total') %>% 
  group_by(geo) %>% summarise(mfd = sum(d))
```

Some countries have problems:

```{r}
kn.diff <- left_join(z.kn.t, z.kn.mf, by='geo') %>%
  mutate (diff = td - mfd) %>%
  filter (diff > 0) 
kable(kn.diff)
```

French is strench (as usual):

```{r}
z.fr.00 <- z %>% filter (age != 'UNK' & geo == 'FR')

z.fr.yr <- z.fr.00 %>% group_by(year, age) %>%
  summarise(value = sum (value, na.rm = T)) %>%
  pivot_wider(names_from = age, values_from = value) 
kable(z.fr.yr)
```

There are no data for age groups for France in 2020. So age-group
analysis for this country is not possible.


## Total excess deaths by age groups

```{r, warning=F, fig.width=pWd, fig.height=pHt}
z <- z %>% filter (age != 'UNK')

z0 <- z %>% filter ( year >= 2015 & year < 2020) %>% 
  group_by(geo, age, week) %>% 
  summarise (d0 = mean(value, na.rm=T))
z1 <- z %>% filter ( year > 2019 ) %>% 
  group_by(geo, age, year, week) %>% 
  summarise ( d1 = sum(value))

zz <- left_join(z0, z1, by=c("week", "age", "geo")) %>% drop_na(d0,d1) %>%
  mutate (exp = (d1 - d0)/d0 * 100, exm = d1 - d0 )
zz <- left_join(zz, nn, by=c('geo'='id'))

## if NA then the country stop reporting in 2020
zz.last.week <- zz %>% filter (year == 2021) %>% 
  group_by (geo, age) %>% 
  summarise (lw = last(week)) %>%
    drop_na(lw)
latestweek <- max (zz.last.week$lw)

### total exmort 
zzt <- zz %>% group_by(geo, age) %>% 
  summarise(country=last(country), exm = sum(exm), nm = sum(d0)) %>%
  mutate (exmp = exm/nm * 100)


zztt <- left_join(zzt, zz.last.week, by=c('geo', 'age')) %>% 
  drop_na(lw) %>% 
  filter (lw >= latestweek - 6) 

exm.total.countries <- zzt %>% filter (age != 'total') %>% summarise (exm = sum(exm))
exm.total.ue <- sum (exm.total.countries$exm)

zzttt <- zzt %>% group_by(age) %>% summarise(exm = sum(exm))  %>%
  mutate (exp = exm/exm.total.ue * 100)

## Wykres słupkowy
source <- "© NI-KW @ github.com/knsm-psw/ | source: Eurostat demo_r_mwk_10 (Deaths by week, sex and 10-year age group)"
note <- sprintf("%s\nFirst column in red = last week reported (excluded countries that reported in %i week or earlier)", source, latestweek - 6 - 1 )

p5 <- zztt %>% ggplot(aes(x = reorder(country, exmp ))) +
  geom_bar(aes(y = exmp), stat="identity", alpha=.4, fill="steelblue" ) +
  geom_text(aes(label=sprintf("%.1f", exmp), y= exmp ), 
            vjust=0.25, hjust=1.25, size=2, color='black' ) +
  geom_text(aes(label=sprintf("%.0f", lw), y= 0 ), 
            vjust=0.25, hjust=-1.25, size=2, color='brown4' ) +
  xlab(label="") +
  ylab(label="") +
  facet_wrap( ~ age, scales = "fixed", ncol = 2, shrink = F) +
  ggtitle(sprintf ("Total Excessive deaths in Europe 2020--2021 (by age groups)"),
          subtitle ="Sum of (deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  theme_nikw() +
  coord_flip() +
  labs(caption=note)

file_out <- sprintf ("EU_exdeaths_total_g6.png")
ggsave(plot=p5, file_out, width=12, height=10)
p5
```

Poland and the rest of Europe:

```{r, warning=F,  fig.width=pWd, fig.height= pHt} 
p6 <- zztt %>%  mutate( base=ifelse(geo=='PL', "1", "0")) %>% 
  ggplot(aes(x = reorder(country, exmp ), fill=as.factor(base))) +
  geom_bar(aes(y = exmp), stat="identity", alpha=.4 ) +
  geom_text(aes(label=sprintf("%.1f", exmp), y= exmp ), 
            vjust=0.25, hjust=1.25, size=2, color='black' ) +
  geom_text(aes(label=sprintf("%.0f", lw), y= 0 ), 
            vjust=0.25, hjust=-1.25, size=2, color='brown4' ) +
  xlab(label="") +
  ylab(label="") +
  ###
  facet_wrap( ~ age, scales = "fixed", ncol = 2, shrink = F) +
  ##coord_cartesian() +
  ggtitle(sprintf("Total Excessive deaths in Europe 2020--2021 (by age groups)"),
          subtitle ="Sum of (deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%") +
  theme_nikw() +
  coord_flip(ylim = c(-12, 45)) +
  scale_fill_manual( values = c( "1"="red", "0"="steelblue" ), guide = FALSE ) +
  labs(caption=note)

#file_out <- sprintf ("EU_exdeaths_total_p_%s.png", this.age)
file_out <- sprintf ("EU_exdeaths_total_p_g6.png")
ggsave(plot=p6, file_out, width=12, height=10)
p6
```

### Dynamics of excess mortality 2020--2021

```{r,warning=F, width=20, height=30}
note.xx <- source
zznn <- zz

## ## ###
p8 <- zznn %>% filter (! geo %in% c('AL', 'AM', 'IE', 'IS', 'UK', 'CY', 'GE', 'LI', 'ME')) %>%
  mutate( base=ifelse(geo=='PL', "1", "0")) %>% 
  mutate(date= as.Date(sprintf("%i-%i-1", year, week), "%Y-%U-%u") ) %>%
  ggplot(aes(x = date, y =exp,  color=as.factor(base ))) +
  ##ggplot(aes(x = date, y =exp)) +
  ##geom_point(size=.1) +
  geom_smooth(aes(x = date, y =exp, group=geo), method="loess", size=.4, se = F, span =.25) +
  geom_hline(yintercept = 50, color="firebrick", alpha=.2, size=1) +
  geom_hline(yintercept = 100, color="red", alpha=.2, size=1) +
  coord_cartesian(ylim = c(-30, 150)) +
  scale_color_manual( values = c( "1"=mainColor, "0"="steelblue" ), guide = FALSE ) +
  geom_hline(yintercept = 0, color='seagreen1', alpha=.25, size=1.5) +
  theme_nikw() +
  facet_wrap( ~ age, scales = "fixed", ncol = 2, shrink = F) +
  ggtitle(sprintf ("Excessive deaths in Europe 2020--2021 (by age groups)"),
          subtitle ="(deaths_2020/2021 - average_2015--2019) / average_2015--2019 * 100%\nLoess smoothing | Poland (pink line) vs other countries (blue lines)") +
  scale_y_continuous(breaks=seq(-25, 150, by=25)) +
  scale_x_date( labels = date_format("%y/%W"), breaks ="8 weeks") +
  xlab(label="year/week") +
  ylab(label="%") +
  labs(caption=note.xx)
p8
file_out <- sprintf ("EU_exdeaths_total_y_g6.png")
ggsave(plot=p8, file_out, width=12, height=10)

```